Composition-climate paper

### 2016-06-19 Composition_Ecology.Rmd
Most text changes complete, minor typos & reformatting.

Forest inventory data are essential for understanding tree species distributions along environmental gradients, and for projecting forest responses to climate change.  However, the stability and representativeness of contemporary tree-climate distributions is poorly understood.  Here we show that historic land use and climate change over the last two centuries have significantly altered tree-climate relationships, using contemporary and historical data for climates and 15 common tree genera in the northern US. Realized niches have shifted towards cooler maximum temperatures, higher minimum temperatures, and higher rainfall. A new attribution method indicates that both historic climate change and land use produced these shifts, with the relative importance varying among genera and climate variables.  The most common climate/land-use interaction is compounding, in which historic land-use exacerbates climatic shifts toward warmer or wetter distributions. Compounding interactions are critical because they suggest that contemporary-based models of species distributions may underestimate species resilience to climate change.

@@

Global change ecologists and biogeographers use contemporary distributions of species, and, increasingly, paleohistorical datasets to map distributions of species and species traits within environmental space REFS. These distributions of species within climate space, combined with future climate scenarios and species distribution models, can be used to assess species exposure and sensitivity to climate change, identify species and habitats of concern, set conservation priorities, and prepare for shifts in habitat suitability caused by climate change  [@pellatt2012using;@iverson2013tree;@iverson1998predicting;@anderson2013nichemodels;@guisan2013predicting]. Forest inventory datasets, such as the US Forest Inventory Analysis (FIA), are particularly rich observational datasets that have, *e.g.* been used to assess early signals of range shifts via the distributions of mature trees and seedlings [@monleon2015evidence;zhu2012failure] and to parameterize models of current and future tree distributions [@iverson2008estimating;@iverson1998predicting].

@@

Over the last several centuries, climate change and land use have both substantially affected species distributions and vegetation structure [@gehrig2007tree] REFS. Understanding the interacting effects of land-use change and climatic change on species-climate relationships is of particular importance, because anthropogenic land use is highly structured and differentially occupies particular habitat types and portions of climate space.  Hence, species distributions and protected natural areas often represent a biased selection from the available environmental space [@scott2001nature].  This biasing can confound studies that use contemporary distributional data to predict species responses to future climate scenarios [@pyke2004habitat].

@@

`r fig_nums("land_use_clim", "*(a-e) Climate change in the Upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions.  Low temperatures in the 19XXs result in lower T~max~ during the modern era, but climate change has resulted in higher T~min~ values than the early-century normals, along with increasing P~ann~.  PDSI shows strong coherence with P~ann~ in this region. (f) Patterns of land-use change in the upper Midwestern United States. Land use data from the NLCD [@jin2013comprehensive].*")`

In the upper Midwestern US, tree distributions and community composition over the last several centuries have been heavily influenced both by the land uses associated with EuroAmerican Settlement in the 1800s [@goring2015preset;@schulte2007homogenization;hanberry2012comparison;@rhemtulla2009historical] and by the climate changes accompanying the end of the Little Ice Age [@hotchkiss2007response;@paquette2013climatic] and current anthropogenic warming [@wicci2011climate].  Modern forests, as represented by the Forest and Inventory Analysis [@woudenberg2010forest], show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System, a survey that covered the western United States prior to widespread land-use conversion in the late 1800s [@goring2015preset;@schulte2007homogenization]. There has been widespread conversion of wildland ecosystems to agricultural and urban land use [@rhemtulla2009historical], particularly along the historical prairie and savanna margins of the region (`r fig_nums("land_use_clim", display = "cite")`f).  Forested regions in the northern Great Lakes States have been also affected by widespread timbering during the late 19th and early 20th-century and ongoing land management REF.  In northern forested regions, we may expect to see expanded ranges of early successional species and reduced abundances of late-successional species, following widespread logging.  Additionally, some tree species, *Ulmus americana* and *Castanea dentata* have experienced rangewide population collapses caused by the introduction of exotic pathogens REFS. These land use changes are superimposed on strong regional gradients of temperature and precipitation in the upper Midwest and two major ecotones, one from prairie/savanna to closed forest, in an approximately west to east direction, and one from conifer dominated, sub-boreal forests to deciduous forest in a north to south direction.

Here we show that the climatic distributions of tree genera in the upper Midwest have significantly shifted over the last 100 years as a result of historic climate change, land use, and pathogen outbreaks, and assess the relative impacts of regional climate shifts and vegetation change in driving these changes.  To this end, we use gridded historic and contemporary climate observations (PRISM) and estimates of pre-settlement and modern forest vegetation [@goring2015preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.  We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates and develop a new method, based on Hellinger distances among tree distributions, to quantify the total change and relative importance of historic vegetation and climate change. We assess whether climate and vegetation changes are compounding, counteracting or confounding based on the magnitude and direction of species-climate shifts. We use this framework to discuss the potential impacts of land-use conversion on species resilience to climate change as predicted by species-distribution models generated from modern distributional data.

@@

`r tab_nums("genera_table", "Tree genera used in the study, along with total spatial coverage (km2)in the pre-settlement and modern eras based on presence within an 8x8km grid cell.  Genera are grouped into broadleaved and needleleaved taxa, then ordered alphabetically.  The total regional coverage is 427,000 km^2^ for the PLS data and 331,000 km^2^ for the FIA data.")`

@@

area_table <- data.frame(Genera = colnames(pls_data)[-1],
                         `PLS Coverage\nkm^2^` = prettyNum(colSums(pls_data > 0) * 64, 
                                                    big.mark = ',', digits = 3)[-1],
                         `FIA Coverage\nkm^2^` = prettyNum(colSums(agg_dens > 0) * 64, 
                                                    big.mark = ',', digits = 3)[-1])

@@

Settlement-era forest data are based on survey data from the Public Land Survey System [PLSS: @white1983history;@schulte2001original;@goring2015preset], aggregated to a 64 km^2 grid across the upper Midwestern United States [@goring2015preset]. We briefly review the methods here. PLSS data originally were collected on a 1 x 1 mi grid across the Upper Midwest from 1830 - 1904 [@schulte2001original;@almendinger1996minnesota;@white1983history]. At each survey, point surveyors noted the closest two to four trees and recorded their common name, distances to the survey point, diameters at breast height and azimuth to the tree. Goring et al. [-@goring2015preset] aggregated this data to an 8x8 km grid, resulting in a gridded data product with an average of 61 survey points per cell, or approximately 120 trees per cell. This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@goring2015preset]. Using the count data produced by Goring et al. [-@goring2015preset], we produce presence-absence distributions for each taxon, across the upper Midwest for 15 common tree genera (Table `r tab_nums("genera_table", display = "cite")`). Common names for certain taxa, such as "Ironwood" can not be clearly resolved to a single tree genera, as such we combine *Ostrya* and *Carpinus* (ironwood), and *Thuja* and *Juniperus* (cedar). Here presence is defined as at least one recorded tree in any 8x8 km grid cell.

Modern forest data are from the US Forest Inventory and Analysis (FIA) program, which began in the 1930s and provide detailed inventories of contemporary tree distributions [@iverson2013tree;@zhu2012failure]. FIA datasets are publicly available surveys of forest composition (for all regions with > 10% forest cover) [@woudenberg2010forest].  FIA uses a nationally standardized sampling procedure, for all land with greater than 10% forest cover.  Each Phase 2/3 FIA plot  consists of four, 7.32m radius, circular sub-plots which are sampled intensively [@woudenberg2010forest].

Comparisons between historical and modern vegetation must address differences in sampling design between the PLSS and FIA datasets [@goring2015preset;@kronenfeld2014validating;@kronenfeld2010influence]. Here the primary tool for maximizing comparability is via aggregation, both by spatially aggregating records from both datasets to a common scale (8x8 km), and taxonomically by aggregating taxa to the level of genera. Nonetheless, differences persist.  For example,  the FIA contains fewer plots per 8 km cell, with more trees per plot than the PLSS data. This leads to higher cell-to-cell variability in the FIA data, but these are not expected to affect the regional-scale analyses here, particularly because we are investigating presence/absence patterns. Additionally, FIA observations are collected only from forested or partially forested landscapes [@woudenberg2010forest], causing the spatial extent of many tree taxa to be lower in the FIA than in the PLSS datasets (`r tab_nums("genera_table", display="cite")`). Hence, scattered trees may be present in regions without permanent or temporary FIA plots, complicating the ability to assign 'absence' within the FIA. However the practice of using FIA data as a surrogate for tree presence/absence data is widespread [@iverson2013tree;@iverson1998predicting], so this complication is a general one faced by forest ecologists and global change ecologists.  In prior comparative analyses of the PLSS and FIA data, Goring et al. [-@goring2015preset] tested the effect of differential sampling design on the mapped patterns of novel and disappearing forests, and found little effect.

@@

Historical and modern climate data are derived from monthly PRISM data for 1895 to 2014 CE [@prism]. The PRISM climate variables are available at an 800m resolution and have been interpolatedfrom climate stationlocations using elevation, aspectand other data.  Estimates for July temperature ($T_{jul}$), January temperature ($T_{jan}$), the annual temperature range ($T_{diff}$), and annual precipitation ($P_{ann}$) were obtained from PRISM and resampled to the 8 x 8km grid used for PLSS and FIA data using bilinear interpolation. Two periods were selected for the calculation of climate normals: 1895 to 1919 for the historical period and 1990 - 2014 for the modern period.  There is an unavoidable temporal mismatch between the historical climate data and the PLSS survey period, on the order of 50 to 70 years.  Few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatio-temporal resolution for this time period.  To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we check the instrumental record against tree-ring-based paleoclimate reconstructions from the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`).

@@

![modelgrid](figures/figureFour_panels.tiff)

`r fig_nums("fourPanel", "(a) Diagram of the 2x2 factorial design used to diagnose the effects of historic climate and vegetation change on the observed shifts in tree-climate relationships. We analyzed observed and hypothetical tree-climate relationships for one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row). (b-e) Representative normalized probability densities for *Larix* for each of the four combinations of climate and vegetation states, plotted relative to Tmax (b), Tdif (c), Tmin (d), and Pann (e).")`

Distributions of tree genera presences along climate gradients are described as probability density functions for both the pre-settlement and modern eras.  Differences in distributions are quantified using Hellinger distances [@matthes2015benchmarking] to compare the probability density functions among the four combined states of vegetation and climate ($V_{H}C_{H}$, $V_{H}C_{M}$, $V_{M}C_{H}$, $V_{M}C_{M}$, where V represents vegetation, C represents climate and the subscripts M and H represent modern and historical conditions, respectively). The total shift ($d_{tot}$) between presettlement conditions and modern conditions is calculated as the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ densities (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right). The effect of historical climate change on shifting climate distributios of tree genera is determined by calculating the Hellinger distance ($d_{C}$) is determined by calculating the Hellinger distance between PLSS and FIA tree distributions when both are overlaid on late 19th-century climates ($V_{H}C_{H}$ versus $V_{M}C_{H}$; `r fig_nums("fourPanel", display = "cite")`a; left column). We develop an attribution index, ($\delta_{d}=d_{c} - d_{l}$), such that negative values of$\delta_{d}$ indicate stronger influence of historical land use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting distributions.

@@

Analysis in this paper is undertaken using R and R Markdown.  All code and raw data is provided in supplemental material and hosted on GitHub (http://github.com/SimonGoring/Composition_Climate).  Analysis makes use of the R packages colorspace, captioner, raster, ggplot2, mgcv, reshape2, plyr, gridExtra, rgdal, RColorBrewer, analogue, and maptools.  Maps use the public domain Natural Earth Data (http://naturalearthdata.com) products for political and landscape-level features.

@@

Regionally, January (minimum) temperatures have increased by `r round(clim_change[4], 1)`^o^C since the 1895-1925 normal `r fig_nums("land_use_clim", display = "cite")`, maximum annual temperature has declined (`r round(clim_change[2], 1)`^o^C), while mean annual temperatures increased by `r round(clim_change[3], 1)`^o^C (not shown). Because of the large shift in minimum temperatures, Tdiff has changed by .  Annual precipitation in the region has increased slightly (`r round(clim_change[1], 1)`mm).  The Palmer Drought Severity Index is closely coherent with Pann trends from 1895 onward and indicate no major climatic trends between 1800 and 1900 (`r fig_nums("land_use_clim", display = "cite")`e), suggesting that early 20^th^ century climates are a reasonable proxy for mid 19^th^ century climates.

@@

![figuretwo](figures/Figure_panelloss.tiff)

`r fig_nums("gain_loss", "Maps showing areas of gain, loss, and continuous presence of tree genera. 'Loss' indicates 8x8km grid cells where PLSS data indicated presence of a genus but FIA data do not record presence (light blue). 'Gain' indicates areas where a genus is absent in the PLSS data but present in the FIA data (red). 'Continuous presence' represents locations where both FIA and PLSS data indicate presence (dark gray). Losses are more common than gains, particularly in the southwestern portion of tree ranges.  Few taxa show gains, although *Populus*, Fraxinus and Tilia show gains of over 10% (Table 2).")`

```{r echo = FALSE}
tab_nums("tab_gain_loss", "Gain and loss of the analyzed tree genera since the pre-settlement era. The first three columns are percentages relative to the total area occupied by the taxon in the presettlement and modern eras. Modern and Historical indicate represent range area (100s of km2) occupied by the tree taxon for the pre-settlement or modern era.  Taxa are sorted by the percent area lost (Column 3).", display = FALSE)
```

For most tree genera, historic range losses dominate over range gains and are spatially structured (Figure 3). Range losses in the Upper Midwest are concentrated in the south and west, along the prairie-savanna-forest ecotone (Figure 3). For *Ulmus* and *Larix*, however, losses occur throughout the range.  Losses are highest for Ostrya/Carpinus and *Larix* when assessed as a proportion of total range (Table 2), and range losses are high generally, with an average of `r round(mean(loss_table$Loss), 1)`% across all taxa, *Larix*, *Ulmus* and *Betula* have the largest absolute losses in range within the Upper Midwest.  Gains, conversely, account for no more than `r round(max(loss_table$Gain),1)`% of any taxon's historical range, with the highest gains observed fo *Populus* (Table 2) and averages `r round(mean(loss_table$Gain),1)`% across all taxa.   Spatial patterns of range gains vary widely among tree genera, and often show less spatial bias than losses (Figure 3).  Range gains for *Picea*, *Larix*, *Abies*, *Tsuga*, *Fagus*, *Ostrya*/*Carpinus*, *Tilia*, and *Fraxinus*, for example, all show a dispersed pattern of gains in a relatively few and widely scattered cells, suggesting that these apparent range gains are mainly caused by sampling uncertainty inherent in local-scale plots and stand heterogeneity, rather than systematic range gains.  Conversely, gains for *Populus* and *Acer*, are concentrated in northern Minnesota, Wisconsin, and Michigan, suggesting range infilling, while *Quercus* gains are consistent with some range extension northwards.  *Pinus*, *Thuja*, and *Ulmus* gains are limited but primarily in the southern portion of their historic range, or beyond it.

@@

![FigureThree](figures/boxplot_clim.tiff)

`r fig_nums("boxplots", "Changes in the realized climate distributions for tree genera in the upper Midwestern United States. In the box-and-whisker plots, the central line indicates the median, box margins indicate 25th and 75th percentiles, whisker tips indicate 5th and 95th percentiles, and dots indicate outliers beyong the 95th percentile. These shifts in realized climate distributions are produced by the joint effects of historical climate change, pathogen-induced tree mortality, and land-use change.")`

The joint effects of historic climate change and land use on tree-climate distributions are illustrated for *Larix* in Figure 2.  For *Larix* we see clear shifts in climate space (from solid to dashed lines) whether modern or presettlement vegetation distributions are used, indicating that historical climate change plays a strong role in shifting the distribution of *Larix* in climate space across all climate variables.  Historical land use also plays a role in shifting distributions, particularly for annual maximum temperature where losses at the highest Tmax values are visible from pre-settlement to modern, however, land use appears toplay a smaller role for the other climate variables.

Shifts in tree-climate distributions are significant for all genera and all climate variables (`r fig_nums("boxplot", display = "cite")`), even accounting for multiple comparisons (all p < 0.0001).  The directionality of shifts between historical and modern climate distributions are consistent across genera: modern maximum annual temperatures are lower for all genera, while minimum temperatures are higher (`r fig_nums("boxplot", display = "cite")`).  Across all genera modern annual precipitation is also higher than in the past.  *Larix* and *Ulmus* show the most extreme changes in climate space overall.  *Ulmus* shows the largest increase in minimum temperature ($\delta_{tmin}$ = 4.2 ^o^C) and the smallest increase occurs for *Larix*. Maximum annual temperature declines most for *Ostrya/Carpinus* ($\delta_{tmax}$ = -1.3 ^o^C) and declines least for *Ulmus* ($\delta_{tmax}$ = -0.4 ^o^C).  *Ulmus* shows the largest increase in mean precipitation ($\delta_{ppt}$ = 86mm), the smallest change is for *Larix* ($\delta_{ppt}$ = 42mm).

@@

```
![hellplot](figures/hellingerplot.tiff)

>>>>

Forest inventory data are essential for understanding tree species distributions along environmental gradients, and for projecting forest responses to climate change.  However, the stability and representativeness of contemporary tree-climate distributions is poorly understood.  Here we show that historic land use and climate change over the last two centuries have significantly altered tree-climate relationships, using contemporary and historical data for climates and 15 common tree genera in the northern US. Realized niches have shifted towards cooler maximum temperatures, higher minimum temperatures, and higher rainfall. A new attribution method indicates that both historic climate change and land use produced these shifts, with the relative importance varying among genera and climate variables.  The most common climate/land-use interaction is compounding, in which historic land-use exacerbates climatic shifts toward warmer and wetter distributions across the historical period. Compounding interactions are critical because they suggest that contemporary-based models of species distributions may underestimate species resilience to climate change.

@@

Global change ecologists and biogeographers use contemporary distributions of species, and, increasingly, paleohistorical datasets to map distributions of species and species traits within environmental space [@moritz2013future;@maguire2015modeling;@ivory2016niche;@pearman2008niche]. Distributions of species within climate space, combined with future climate scenarios and species distribution models, can be used to assess species exposure and sensitivity to climate change, identify species and habitats of concern, set conservation priorities, and prepare for shifts in habitat suitability caused by climate change  [@pellatt2012using;@iverson2013tree;@iverson1998predicting;@anderson2013nichemodels;@guisan2013predicting]. Forest inventory datasets, such as the US Forest Inventory Analysis (FIA), are particularly rich observational datasets that have, *e.g.* been used to assess early signals of range shifts via the distributions of mature trees and seedlings [@monleon2015evidence;@zhu2012failure] and to parameterize models of current and future tree distributions [@iverson2008estimating;@iverson1998predicting].

@@

Over the last several centuries, climate change and land use have both substantially affected species distributions and vegetation structure [*e.g.*, @gehrig2007tree]. Understanding the interacting effects of land-use change and climatic change on species-climate relationships is of particular importance, because anthropogenic land use is highly structured and differentially occupies particular habitat types and portions of climate space.  Hence, species distributions and protected natural areas often represent a biased selection from the available environmental space [@scott2001nature].  This biasing can confound studies that use contemporary distributional data to predict species responses to future climate scenarios [@pyke2004habitat].

@@

`r fig_nums("land_use_clim", "*(a-e) Climate change in the Upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions.  Regionally low temperatures in the 1990s result in lower $T_{max}$ during the modern era, but climate change has resulted in higher $T_{min}$ values than the early-century normals, along with increasing $P_{ann}$.  PDSI shows strong coherence with $P_{ann}$ in this region. (f) Patterns of land-use change in the upper Midwestern United States. Land use data from the NLCD [@jin2013comprehensive].*")`

In the upper Midwestern US, tree distributions and community composition over the last several centuries have been heavily influenced both by the land uses associated with EuroAmerican Settlement in the 1800s [@gorng2016preset;@schulte2007homogenization;hanberry2012comparison;@rhemtulla2009historical] and by the climate changes accompanying the end of the Little Ice Age [@hotchkiss2007response;@paquette2013climatic] and current anthropogenic warming [@wicci2011climate].  Modern forests, as represented by the Forest and Inventory Analysis [@woudenberg2010forest], show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System, a survey that covered the western United States prior to widespread land-use conversion in the late 1800s [@gorng2016preset;@schulte2007homogenization]. There has been widespread conversion of wildland ecosystems to agricultural and urban land use [@rhemtulla2009historical], particularly along the historical prairie and savanna margins of the region (`r fig_nums("land_use_clim", display = "cite")`f).  Forested regions in the northern Great Lakes States have been also affected by widespread timbering during the late 19th and early 20th-century and ongoing land management REF.  In northern forested regions, we may expect to see expanded ranges of early successional species and reduced abundances of late-successional species, following widespread logging.  Additionally, some tree species, *Ulmus americana* and *Castanea dentata* have experienced rangewide population collapses caused by the introduction of exotic pathogens REFS. These land use changes are superimposed on strong regional gradients of temperature and precipitation in the upper Midwest and two major ecotones, one from prairie/savanna to closed forest, in an approximately west to east direction, and one from conifer dominated, sub-boreal forests to deciduous forest in a north to south direction.

Here we show that the climatic distributions of tree genera in the upper Midwest have significantly shifted over the last 100 years as a result of historic climate change, land use, and pathogen outbreaks, and we assess the relative impacts of regional climate shifts and vegetation change in driving these changes.  To this end, we use gridded historic and contemporary climate observations (PRISM) and estimates of pre-settlement and modern forest vegetation [@gorng2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.  We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates and develop a new method, based on Hellinger distances among tree distributions, to quantify the total change and relative importance of historic vegetation and climate change. We assess whether climate and vegetation changes are compounding, counteracting or confounding based on the magnitude and direction of species-climate shifts. We use this framework to discuss the potential impacts of land-use conversion on species resilience to climate change as predicted by species-distribution models generated from modern distributional data.

@@

`r tab_nums("genera_table", "Tree genera used in the study, along with total spatial coverage (km^2^)in the pre-settlement and modern eras based on presence within an 8x8km grid cell.  Genera are grouped into broadleaved and needleleaved taxa, then ordered alphabetically.  The total regional coverage is 427,000 km^2^ for the PLS data and 331,000 km^2^ for the FIA data.")`

@@

area_table <- data.frame(Genera = colnames(pls_data)[-1],
                         `PLS\\sCoverage\\nkm^2^` = prettyNum(colSums(pls_data > 0) * 64, 
                                                    big.mark = ',', digits = 3)[-1],
                         `FIA\\sCoverage\\nkm^2^` = prettyNum(colSums(agg_dens > 0) * 64, 
                                                    big.mark = ',', digits = 3)[-1])

colnames(area_table)[2] <- "PLS\\sCoverage\\nkm^2^"
colnames(area_table)[3] <- "FIA\\sCoverage\\nkm^2^"

@@

Settlement-era forest data are based on survey data from the Public Land Survey System [PLS: @white1983history;@schulte2001original;@gorng2016preset], aggregated to a 64 km^2 grid across the upper Midwestern United States [@gorng2016preset]. We briefly review the methods here. PLS data originally were collected on a 1 x 1 mi grid across the Upper Midwest from 1830 - 1904 [@schulte2001original;@almendinger1996minnesota;@white1983history]. At each survey, point surveyors noted the closest two to four trees and recorded their common name, distances to the survey point, diameters at breast height and azimuth to the tree. Goring et al. [-@gorng2016preset] aggregated this data to an 8x8 km grid, resulting in a gridded data product with an average of 61 survey points per cell, or approximately 120 trees per cell. This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@gorng2016preset]. Using the count data produced by Goring et al. [-@gorng2016preset], we produce presence-absence distributions for each taxon, across the upper Midwest for 15 common tree genera (`r tab_nums("genera_table", display = "cite")`). Common names for certain taxa, such as "Ironwood" can not be clearly resolved to a single tree genera, as such we combine *Ostrya* and *Carpinus* (ironwood), and *Thuja* and *Juniperus* (cedar). Here presence is defined as at least one recorded tree in any 8x8 km grid cell.

Modern forest data are from the US Forest Inventory and Analysis (FIA) program, which began in the 1930s and provide detailed inventories of contemporary tree distributions [@iverson2013tree;@zhu2012failure]. FIA datasets are publicly available surveys of forest composition across the United States [@woudenberg2010forest].  FIA uses a nationally standardized sampling procedure, for all land with greater than 10% forest cover.  Each Phase 2 and 3 FIA plot  consists of four, 7.32m radius, circular sub-plots which are sampled intensively [@woudenberg2010forest].

Comparisons between historical and modern vegetation must address differences in sampling design between the PLS and FIA datasets [@gorng2016preset;@kronenfeld2014validating;@kronenfeld2010influence]. Here the primary tool for maximizing comparability is via aggregation, both by spatially aggregating records from both datasets to a common scale (8x8 km), and taxonomically by aggregating taxa to the level of genera. Nonetheless, differences persist.  For example,  the FIA contains fewer plots per 8 km cell, with more trees per plot than the PLS data. This leads to higher cell-to-cell variability in the FIA data, but these are not expected to affect the regional-scale analyses here, particularly because we are investigating presence-absence patterns. Additionally, FIA observations are collected only from forested or partially forested landscapes [@woudenberg2010forest], causing the spatial extent of many tree taxa to be lower in the FIA than in the PLS datasets (`r tab_nums("genera_table", display="cite")`). Hence, scattered trees may be present in regions without permanent or temporary FIA plots, complicating the ability to assign 'absence' within the FIA. However the practice of using FIA data as a surrogate for tree presence-absence data is widespread [@iverson2013tree;@iverson1998predicting], so this complication is a general one faced by forest ecologists and global change ecologists.  In prior comparative analyses of the PLS and FIA data, Goring et al. [-@gorng2016preset] tested the effect of differential sampling design on the mapped patterns of novel and disappearing forests, and found little effect.

@@

Historical and modern climate data are derived from monthly PRISM data for 1895 to 2014 CE [@prism2004]. The PRISM climate variables are available at an 800m resolution and have been interpolatedfrom climate stationlocations using elevation, aspectand other data.  Estimates for July temperature ($T_{jul}$), January temperature ($T_{jan}$), the annual temperature range ($T_{diff}$), and annual precipitation ($P_{ann}$) were obtained from PRISM and resampled to the 8 x 8km grid used for PLS and FIA data using bilinear interpolation. Two periods were selected for the calculation of climate normals: 1895 to 1919 for the historical period and 1990 - 2014 for the modern period.  There is an unavoidable temporal mismatch between the historical climate data and the PLS survey period, on the order of 50 to 70 years.  Few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatio-temporal resolution for this time period, although recent methodological advances [@tiptoninreview] may provide estimates from oberver datasets as far back as 1825.  To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we check the instrumental record against tree-ring-based paleoclimate reconstructions from the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`).

@@

![](figures/figureFour_panels.tiff)

`r fig_nums("fourPanel", "(a) Diagram of the 2x2 factorial design used to diagnose the effects of historic climate and vegetation change on the observed shifts in tree-climate relationships. We analyzed observed and hypothetical tree-climate relationships for one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row). (b-e) Representative normalized probability densities for *Larix* for each of the four combinations of climate and vegetation states, plotted relative to $T_{max}$ (b), $T_{diff}$ (c), $T_{min}$ (d), and $P_{ann}$ (e).")`

Distributions of tree genera presences along climate gradients are described as probability density functions for both the pre-settlement and modern eras.  Differences in distributions are quantified using Hellinger distances [@matthes2016benchmarking] to compare the probability density functions among the four combined states of vegetation and climate ($V_{H}C_{H}$, $V_{H}C_{M}$, $V_{M}C_{H}$, $V_{M}C_{M}$, where V represents vegetation, C represents climate and the subscripts M and H represent modern and historical conditions, respectively). The total shift ($d_{tot}$) between presettlement conditions and modern conditions is calculated as the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ densities (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right). The effect of historical climate change on shifting climate distributios of tree genera is determined by calculating the Hellinger distance ($d_{C}$) between PLS and FIA tree distributions when both are overlaid on late 19th-century climates ($V_{H}C_{H}$ versus $V_{M}C_{H}$; `r fig_nums("fourPanel", display = "cite")`a; left column). We develop an attribution index, ($\delta_{d}=d_{c} - d_{l}$), such that negative values of $\delta_{d}$ indicate stronger influence of historical land use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting distributions.

@@

Analysis in this paper is undertaken using R [@Team2014] and R Markdown [@rmarkdown2016].  All code and raw data is provided in supplemental material and hosted on GitHub (http://github.com/SimonGoring/Composition_Climate).  Analysis makes use of the R packages colorspace, captioner, raster, ggplot2, mgcv, reshape2, plyr, gridExtra, rgdal, RColorBrewer, analogue, and maptools.  Maps use the public domain Natural Earth Data (http://naturalearthdata.com) products for political and landscape-level features.

@@

Regionally, $T_{min}$ has increased by `r round(clim_change[4], 1)`^o^C since the 1895-1924 normal `r fig_nums("land_use_clim", display = "cite")`, $T_{max}$ has declined (`r round(clim_change[2], 1)`^o^C).  Because of the large shift in minimum temperatures, $T_{diff}$ decreased by `r round(clim_change[3], 1)`^o^C.  Annual precipitation in the region has increased slightly (`r round(clim_change[1], 1)`mm).  The Palmer Drought Severity Index is closely coherent with $P_{ann}$ trends from 1895 onward and indicate no major climatic trends between 1800 and 1900 (`r fig_nums("land_use_clim", display = "cite")`e), suggesting that early 20^th^ century climates are a reasonable proxy for mid 19^th^ century climates.

@@

![](figures/Figure_panelloss.tiff)

`r fig_nums("gain_loss", "Maps showing areas of gain, loss, and continuous presence of tree genera. 'Loss' indicates 8x8km grid cells where PLS data indicated presence of a genus but FIA data do not record presence (light blue). 'Gain' indicates areas where a genus is absent in the PLS data but present in the FIA data (red). 'Continuous presence' represents locations where both FIA and PLS data indicate presence (dark gray). Losses are more common than gains, particularly in the southwestern portion of tree ranges.  Few taxa show gains, although *Populus*, Fraxinus and Tilia show gains of over 10% (Table 2).")`

```{r echo = FALSE}
tab_nums("tab_gain_loss", "Gain and loss of the analyzed tree genera since the pre-settlement era. The first three columns are percentages relative to the total area occupied by the taxon in the presettlement and modern eras. Modern and Historical indicate represent range area (100s of km^2^) occupied by the tree taxon for the pre-settlement or modern era.  Taxa are sorted by the percent area lost (Column 3).", display = FALSE)
```

For most tree genera, historic range losses dominate over range gains and are spatially structured (Figure 3). Range losses in the Upper Midwest are concentrated in the south and west, along the prairie-savanna-forest ecotone (Figure 3). For *Ulmus* and *Larix*, however, losses occur throughout the range.  Losses are highest for *Ostrya*/*Carpinus* and *Larix* when assessed as a proportion of total range (Table 2), and range losses are high generally, with an average of `r round(mean(loss_table$Loss), 1)`% across all taxa, *Larix*, *Ulmus* and *Betula* have the largest absolute losses in range within the Upper Midwest.  Gains, conversely, account for no more than `r round(max(loss_table$Gain),1)`% of any taxon's historical range, with the highest gains observed for *Populus* (Table 2) and averages `r round(mean(loss_table$Gain),1)`% across all taxa.   Spatial patterns of range gains vary widely among tree genera, and often show less spatial bias than losses (Figure 3).  Range gains for *Picea*, *Larix*, *Abies*, *Tsuga*, *Fagus*, *Ostrya*/*Carpinus*, *Tilia*, and *Fraxinus*, for example, all show a dispersed pattern of gains in a relatively few and widely scattered cells, suggesting that these apparent range gains are mainly caused by sampling uncertainty inherent in local-scale plots and stand heterogeneity, rather than systematic range gains.  Conversely, gains for *Populus* and *Acer*, are concentrated in northern Minnesota, Wisconsin, and Michigan, suggesting range infilling, while *Quercus* gains are consistent with some range extension northwards.  *Pinus*, *Thuja*, and *Ulmus* gains are limited but primarily in the southern portion of their historic range, or beyond it.

@@

![](figures/boxplot_clim.tiff)

`r fig_nums("boxplots", "Changes in the realized climate distributions for tree genera in the upper Midwestern United States. In the box-and-whisker plots, the central line indicates the median, box margins indicate 25^th^ and 75^th^ percentiles, whisker tips indicate 5^th^ and 95^th^ percentiles, and dots indicate outliers beyong the 95^th^ percentile. These shifts in realized climate distributions are produced by the joint effects of historical climate change, pathogen-induced tree mortality, and land-use change.")`

The joint effects of historic climate change and land use on tree-climate distributions are illustrated for *Larix* in Figure 2.  For *Larix* we see clear shifts in climate space (from solid to dashed lines) whether modern or presettlement vegetation distributions are used, indicating that historical climate change plays a strong role in shifting the distribution of *Larix* in climate space across all climate variables.  Historical land use also plays a role in shifting distributions, particularly for $T_{max}$, where losses at the highest $T_{max}$ values are visible from pre-settlement to modern, however, land use appears to play a smaller role for the other climate variables.

Shifts in tree-climate distributions are significant for all genera and all climate variables (`r fig_nums("boxplot", display = "cite")`), even accounting for multiple comparisons (all p < 0.0001).  The directionality of shifts between historical and modern climate distributions are consistent across genera: modern $T_{max}$ is lower for all genera, while $T_{min}$ is higher (`r fig_nums("boxplot", display = "cite")`).  Across all genera modern annual precipitation is also higher than in the past.  *Larix* and *Ulmus* show the most extreme changes in climate space overall.  *Ulmus* shows the largest increase in minimum temperature ($\delta_{Tmin}$ = 4.2 ^o^C) and the smallest increase occurs for *Larix*. Maximum annual temperature declines most for *Ostrya/Carpinus* ($\delta_{Tmax}$ = -1.3 ^o^C) and declines least for *Ulmus* ($\delta_{Tmax}$ = -0.4 ^o^C).  *Ulmus* shows the largest increase in mean precipitation ($\delta_{Pann}$ = 86mm), the smallest change is for *Larix* ($\delta_{Pann}$ = 42mm).

@@

```
![](figures/hellingerplot.tiff)
